---
title: "The callers: one example each"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The callers: one example each}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_chunk$set(eval = requireNamespace("simcross", quietly = TRUE))
set.seed(1)
```

nilHMM ships six named callers. They all share the 3-state REF/HET/ALT chain and
the breeding-design priors, and all return the same segment schema — so you can
run several on the same data and compare directly. They differ in the **emission**
model, the **duration** prior, and the **input** they expect. This vignette runs
one minimal example of each on self-contained synthetic data.

```{r setup}
library(nilHMM)
```

## Synthetic data

We build one small BC2S2 cohort (donor **B** crossed onto a recurrent parent
**A**) with [**simcross**](https://cran.r-project.org/package=simcross) (real
meiosis and recombination on a genetic map), then layer allelic read counts.
The result carries **both** the counts (`n_ref`/`n_alt`, for the count callers) and
a hard genotype `g` in `{0,1,2,3}` (for the genotype callers), from the same
simulated genotypes.

```{r sim}
library(simcross)

sim_cohort <- function(n = 8L, m = 150L, n_chr = 2L, L = 100,
                       nbc = 2L, nself = 2L, depth = 6, err = 0.01, miss = 0.05) {
  rec <- create_parent(L, 1); don <- create_parent(L, 2)
  one_line <- function() {
    ind <- cross(rec, don)                                # F1
    for (b in seq_len(nbc))   ind <- cross(ind, rec)      # backcross to recurrent
    for (s in seq_len(nself)) ind <- cross(ind, ind)      # selfing
    ind
  }
  map <- seq(0, L, length.out = m)
  do.call(rbind, lapply(seq_len(n_chr), function(ch) {
    g <- as.vector(convert2geno(lapply(seq_len(n), function(i) one_line()), map)) - 1L
    N <- length(g); p_alt <- c(err, 0.5, 1 - err)[g + 1L]
    d <- rpois(N, depth); na <- rbinom(N, d, p_alt)
    g_obs <- g; g_obs[d == 0L | runif(N) < miss] <- 3L
    data.frame(name = rep(sprintf("NIL%02d", seq_len(n)), times = m),
               donor = "B", family = "B", chr = ch,
               pos = rep(as.integer(map * 1e6), each = n),
               n_ref = d - na, n_alt = na, g = as.integer(g_obs),
               truth = as.integer(g))                       # noise-free simcross state
  }))
}
d <- sim_cohort()   # donor B (ALT) on recurrent A (REF)
counts <- d[, c("name", "donor", "chr", "pos", "n_ref", "n_alt")]
gt     <- d[, c("name", "donor", "chr", "pos", "g")]
```

## `nnil` — count and genotype

Holland's nNIL caller. With allelic counts it pools single-read observations along
a segment via a BetaBinomial emission (`err`, `conc`; `fit_means = TRUE` EM-fits
the per-state alt fractions). Given only a called genotype `g`, it switches to the
categorical genotype emission automatically.

```{r nnil}
nnil_count <- call_ancestry(counts, caller = "nnil", design = "BC2S2",
                            rrate = 1e-4, err = 0.01)
nnil_gt    <- call_ancestry(gt, caller = "nnil", design = "BC2S2")  # g-only -> gt emission
nrow(nnil_count); nrow(nnil_gt)
```

## `rtiger` — rigidity segmentation

A Julia-free port of the RTIGER rigidity HMM (EM + Viterbi + border re-placement).
`rigidity` is the integer minimum run length.

```{r rtiger}
rt <- call_ancestry(counts, caller = "rtiger", design = "BC2S2",
                    rigidity = 5L, seed = 1L)
head(rt, 3)
```

## `binhmm` — per-bin calling

Bins the genome (default 1 Mb) and calls per-bin state with an anchored 3-state
Gaussian-emission HMM. Good for noisy or uneven coverage.

```{r binhmm}
bh <- call_ancestry(counts, caller = "binhmm", design = "BC2S2", bin_size = 5e6)
head(bh, 3)
```

## `atlas` — competitive-alignment (GOOGA)

For transcript / competitive-alignment data: `n_ref`/`n_alt` are the recurrent and
donor read counts, thresholded into genotype calls (GOOGA style) then smoothed.

```{r atlas}
at <- call_ancestry(counts, caller = "atlas", design = "BC2S2",
                    atlas_thresh = 0.95, atlas_het = 0.25, atlas_min_reads = 5L)
head(at, 3)
```

## `lbimpute` — very low coverage

A native port of LB-Impute (Fragoso et al. 2014) for `<1×` biallelic populations: a
coverage-aware emission bounded by `genotypeerr`, and a distance-dependent
transition (recombination scales with the marker gap over `recombdist`). No design
priors needed.

```{r lbimpute}
lb <- call_ancestry(counts, caller = "lbimpute", recombdist = 1e7, genotypeerr = 0.05)
head(lb, 3)
```

## `fsfhap` — full-sib families

A port of TASSEL's FSFHap. Unlike the per-line callers it **pools each family**, so
it needs a `family` grouping and a called genotype `g`. See `vignette("fsfhap")`
for the full workflow (HapMap + pedigree input, design routing, parallelism).

```{r fsfhap}
fam <- d[, c("name", "family", "chr", "pos", "g")]
# (a real family needs enough segregating markers; see vignette("fsfhap"))
```

## Comparing callers

Because the output schema is shared, comparison is a table join away — e.g. how
many segments and which states each caller produced on the same counts:

```{r compare}
summ <- function(x) c(segments = nrow(x), states = length(unique(x$state)))
rbind(nnil = summ(nnil_count), rtiger = summ(rt), binhmm = summ(bh),
      atlas = summ(at), lbimpute = summ(lb))
```

### Chromosome painting

The real payoff of the shared schema is that you can **paint every caller on the
same axes** — against the simcross **ground truth** — and eyeball whether the
independent methods recover the donor (**B**) blocks. Each row is a NIL, each
column a chromosome, and within a cell the tracks are stacked as bands
(REF gold / HET green / ALT purple), with the noise-free truth on top.

The truth is just `to_segments()` on the simulated genotype *before* the depth
and missingness were added (the `truth` column). `paint_calls()` then stacks it
above the callers: `rbind` each track's segments with a column naming it, and
pass that column as `track` (its levels stack top-down, so truth goes first).

```{r paint, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("simcross", quietly = TRUE), fig.width = 8, fig.height = 5.5}
tracks <- list("nnil (count)" = nnil_count, "nnil (gt)" = nnil_gt,
               rtiger = rt, binhmm = bh, atlas = at, lbimpute = lb)
comparison <- do.call(rbind, Map(function(seg, m) { seg$method <- m; seg },
                                 tracks, names(tracks)))

# simcross ground truth (noise-free) as the top track
truth <- to_segments(data.frame(name = d$name, donor = d$donor,
                                 chr = d$chr, pos = d$pos, state = d$truth))
truth$method <- "truth (simcross)"

comparison <- rbind(truth, comparison)
comparison$method <- factor(comparison$method,
                            levels = c("truth (simcross)", names(tracks)))
paint_calls(comparison, track = "method", samples = sprintf("NIL%02d", 1:4))
```

The callers recover the true donor blocks — each caller's band lines up with the
truth on top — with expected method-specific behaviour (e.g. `binhmm` paints
broader per-bin blocks). This is the same painting used for the real
coverage-sweep NILs, where the truth track is instead an independent data source
or a high-confidence call set.

For per-caller parameters and lineage, see `?call_ancestry` and the README caller
table.
```
