---
title: "The callers: one example each"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The callers: one example each}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_chunk$set(eval = requireNamespace("simcross", quietly = TRUE))
set.seed(1)
```

nilHMM ships eleven named callers. The HMM callers share the 3-state REF/HET/ALT
chain and the breeding-design priors (the two no-HMM baselines `ml`/`hwemap` use a
flat / Hardy–Weinberg prior instead), and all return the same segment schema — so you can
run several on the same data and compare directly. They differ in the **emission**
model, the **duration** prior, and the **input** they expect. This vignette runs
one minimal example of each on self-contained synthetic data.

```{r setup}
library(nilHMM)
```

## Synthetic data

`simulate_nil()` builds a BC2S2 cohort (donor **B** on recurrent parent **A**) on
the bundled maize map — the **truth** — and `simulate_counts()` degrades it to
observed data carrying **both** allelic counts (`n_ref`/`n_alt`, for the count
callers) and a hard genotype `g` in `{0,1,2,3}` (for the genotype callers):

```{r sim}
truth <- simulate_nil("BC2S2", n = 8, chr = 1:2, n_markers = 300, donor = "B",
                      names = sprintf("NIL%02d", 1:8), seed = 1)
obs   <- simulate_counts(truth, depth = 6, seed = 1)
counts <- obs[c("name", "donor", "chr", "pos", "n_ref", "n_alt")]
gt     <- obs[c("name", "donor", "chr", "pos", "g")]
```

## The four grid callers (emission × duration)

The pure cells of the engine's grid share the REF/HET/ALT chain and differ only in
two axes: the emission (categorical `gt` vs count/BetaBinomial) and the duration
(geometric vs rigidity).

* **`nnil`** — Holland's nNIL: categorical `gt` emission + geometric duration. A
  called genotype `g` is used directly; read counts are first hard-called
  (1/3–2/3 cutoffs). Error model: `germ`, `gert`, `p`, `mr`, `nir`.
* **`bbnil`** — the low-coverage count extension: count/BetaBinomial emission
  (`err`, `conc`; `fit_means = TRUE` EM-fits the per-state alt fractions) + the
  same geometric duration.
* **`catiger`** — categorical `gt` emission + rigidity duration.
* **`rtiger`** — count/BetaBinomial emission + rigidity duration (below).

```{r grid}
nnil    <- call_ancestry(gt,     caller = "nnil",    design = "BC2S2")           # gt + geometric
bbnil   <- call_ancestry(counts, caller = "bbnil",   design = "BC2S2",
                         rrate = 1e-4, err = 0.01)                               # count + geometric
catiger <- call_ancestry(gt,     caller = "catiger", design = "BC2S2", rigidity = 5L)  # gt + rigidity
nrow(nnil); nrow(bbnil); nrow(catiger)
```

## `rtiger` — rigidity segmentation

A Julia-free port of the RTIGER rigidity HMM (EM + Viterbi + border re-placement),
the count-emission rigidity cell. `rigidity` is the integer minimum run length.

```{r rtiger}
rt <- call_ancestry(counts, caller = "rtiger", design = "BC2S2",
                    rigidity = 5L, seed = 1L)
head(rt, 3)
```

## `binhmm` — per-bin calling

Bins the genome (default 1 Mb) and calls per-bin state with an anchored 3-state
Gaussian-emission HMM. Good for noisy or uneven coverage.

```{r binhmm}
bh <- call_ancestry(counts, caller = "binhmm", design = "BC2S2", bin_size = 5e6)
head(bh, 3)
```

## `googa` / `atlas` — competitive-alignment (transcript ancestry)

For transcript / competitive-alignment data: `n_ref`/`n_alt` are the recurrent and
donor read counts, thresholded into hard genotype calls (`atlas_thresh`,
`atlas_het`, `atlas_min_reads`) then smoothed. **`googa`** is the faithful
reproduction — gt + **geometric**, matching GOOGA's recombination-fraction F2 HMM
(Flagel 2019; Veltsos 2024), which carries no rigidity. **`atlas`** is this work's
transcript caller: the same thresholding with the **rigidity** duration.

```{r atlas}
gg <- call_ancestry(counts, caller = "googa", design = "BC2S2",
                    atlas_thresh = 0.95, atlas_het = 0.25, atlas_min_reads = 5L)
at <- call_ancestry(counts, caller = "atlas", design = "BC2S2", rigidity = 5L,
                    atlas_thresh = 0.95, atlas_het = 0.25, atlas_min_reads = 5L)
head(gg, 3); head(at, 3)
```

## `lbimpute` — very low coverage

A native port of LB-Impute (Fragoso et al. 2014) for `<1×` biallelic populations: a
coverage-aware emission bounded by `genotypeerr`, and a distance-dependent
transition (recombination scales with the marker gap over `recombdist`). No design
priors needed.

```{r lbimpute}
lb <- call_ancestry(counts, caller = "lbimpute", recombdist = 1e7, genotypeerr = 0.05)
head(lb, 3)
```

## `ml` / `hwemap` — no-HMM per-site baselines

Two per-site genotype callers with **no** linkage/duration model — each
`(marker, sample)` is decided independently from its own read counts (backed by
`call_gt()`). `ml` uses a flat prior (pure argmax genotype-likelihood = maximum
likelihood; het-blind at depth 1). `hwemap` uses the per-marker Hardy–Weinberg
prior (posterior MAP; deliberately het-**excess** at low depth) — the reference
the HMM callers must beat. These are not to be confused with the deterministic
`interpolate_genotype()` densifier.

```{r baselines}
ml_calls <- call_ancestry(counts, caller = "ml",     design = "BC2S2")
hw_calls <- call_ancestry(counts, caller = "hwemap", design = "BC2S2")
nrow(ml_calls); nrow(hw_calls)
```

## `fsfhap` — full-sib families

A port of TASSEL's FSFHap. Unlike the per-line callers it **pools each family**, so
it needs a `family` grouping and a called genotype `g`. See `vignette("fsfhap")`
for the full workflow (HapMap + pedigree input, design routing, parallelism).

```{r fsfhap}
fam <- transform(gt, family = donor)          # a family grouping on the genotype table
# (a real family needs enough segregating markers; see vignette("fsfhap"))
```

## Comparing callers

Because the output schema is shared, comparison is a table join away — e.g. how
many segments and which states each caller produced on the same counts:

```{r compare}
summ <- function(x) c(segments = nrow(x), states = length(unique(x$state)))
rbind(nnil = summ(nnil), bbnil = summ(bbnil), catiger = summ(catiger),
      rtiger = summ(rt), binhmm = summ(bh), googa = summ(gg), atlas = summ(at),
      lbimpute = summ(lb), ml = summ(ml_calls), hwemap = summ(hw_calls))
```

### Chromosome painting

The real payoff of the shared schema is that you can **paint every caller on the
same axes** — against the simcross **ground truth** — and eyeball whether the
independent methods recover the donor (**B**) blocks. Each row is a NIL, each
column a chromosome, and within a cell the tracks are stacked as bands
(REF gold / HET green / ALT purple), with the noise-free truth on top.

The truth track is just `to_segments()` on the `simulate_nil()` table (its `state`
is the noise-free donor mosaic, before `simulate_counts()` added depth and
missingness). `paint_calls()` then stacks it above the callers: `rbind` each
track's segments with a column naming it, and pass that column as `track` (its
levels stack top-down, so truth goes first).

```{r paint, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("simcross", quietly = TRUE), fig.width = 8, fig.height = 5.5}
tracks <- list("nnil (gt)" = nnil, "bbnil (count)" = bbnil, catiger = catiger,
               rtiger = rt, binhmm = bh, googa = gg, atlas = at, lbimpute = lb,
               ml = ml_calls, hwemap = hw_calls)
comparison <- do.call(rbind, Map(function(seg, m) { seg$method <- m; seg },
                                 tracks, names(tracks)))

# simcross ground truth (the simulate_nil() table) as the top track
truth_seg <- to_segments(truth)
truth_seg$method <- "truth (simcross)"

comparison <- rbind(truth_seg, comparison)
comparison$method <- factor(comparison$method,
                            levels = c("truth (simcross)", names(tracks)))
paint_calls(comparison, track = "method", samples = sprintf("NIL%02d", 1:4))
```

The callers recover the true donor blocks — each caller's band lines up with the
truth on top — with expected method-specific behaviour (e.g. `binhmm` paints
broader per-bin blocks). This is the same painting used for the real
coverage-sweep NILs, where the truth track is instead an independent data source
or a high-confidence call set.

For per-caller parameters and lineage, see `?call_ancestry` and the README caller
table.
```
