---
title: "The engine: one chain, two swappable axes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The engine: one chain, two swappable axes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_chunk$set(eval = requireNamespace("simcross", quietly = TRUE))
set.seed(7)
```

`call_ancestry()` is a thin convenience layer. Underneath, every count/geometric
caller is the **same 3-state (REF/HET/ALT) HMM** parameterised along two
independent axes plus a set of design priors:

```
caller  =  emission model  ×  duration prior  +  design priors
```

- **Emission** — how an observation maps to a state likelihood:
  `emission_count()` (BetaBinomial over ref/alt depths) or `emission_gt()`
  (categorical genotype-error model).
- **Duration** — the transition prior on run lengths:
  `duration_geometric()`, `duration_rigidity()` (minimum run length), or
  `duration_hsmm()`.
- **Design priors** — `breeding_prior()` turns a breeding design into the
  genotype-frequency prior `c(REF, HET, ALT)`; the engine uses its `HET`/`ALT`
  as the state frequencies `f_1`/`f_2`.

This vignette drives the engine directly with `fit()` → `decode()` →
`to_segments()`, which is exactly what `call_ancestry()` chains for you.

```{r setup}
library(nilHMM)
```

## The building blocks

Each axis is a small constructor returning a spec:

```{r blocks}
breeding_prior("BC2S2")    # genotype-freq prior c(REF, HET, ALT) from the design
emission_count(err = 0.01, conc = 20)
duration_geometric(rrate = 1e-4)
```

## Fit and decode by hand

`fit()` takes a per-(sample, chromosome) observation frame — `n` = depth, `a` =
alt-read count — an emission spec, a duration spec, and priors, and returns a
fitted model. `decode()` runs Viterbi and returns one **state per observation**
(coordinate-free: `0/1/2`).

We use `simulate_nil()` (a BC2S2 backcross of a donor **B** onto a recurrent
parent **A**, on chromosome 1 of the bundled map) and `simulate_counts()`, so the
observations come from a real recombination track:

```{r fit-decode}
# a BC2S2 cohort on chromosome 1; take the line with the most donor as our example
truth <- simulate_nil("BC2S2", n = 12, chr = 1, n_markers = 24, donor = "B", seed = 7)
line  <- truth[truth$name == names(which.max(tapply(truth$state == 2L, truth$name, mean))), ]
obs   <- simulate_counts(line, depth = 8, seed = 7)
o     <- data.frame(n = obs$n_ref + obs$n_alt, a = obs$n_alt)    # fit()/decode(): n = depth, a = alt

bp     <- breeding_prior("BC2S2")                                # c(REF, HET, ALT) from the design
priors <- list(f_1 = bp[["HET"]], f_2 = bp[["ALT"]])             # the engine's state frequencies
model  <- fit(o, emission_count(err = 0.01), duration_geometric(1e-4), priors = priors)
rbind(simulated = line$state, decoded = decode(model, o))        # decode recovers the track
```

## Coordinates back on: `to_segments()`

`decode()` is coordinate-free. Attach `name/chr/pos` and run `to_segments()` to
collapse the per-marker states into the common segment schema (this is the second
half of `call_ancestry()`):

```{r segments}
states <- data.frame(name = "S1", donor = "B", chr = line$chr, pos = line$pos,
                     state = decode(model, o))
to_segments(states)
```

## Swapping an axis

Because the axes are independent, changing the caller's behaviour is a matter of
swapping one spec. Keep the same count emission but replace the geometric duration
with a **rigidity** prior (minimum run length 4) — the RTIGER-style segmentation.
Rigidity enforces a minimum run length, so it resists short state changes; on this
clean, well-covered line the call is essentially unchanged, but on noisy
low-coverage data it suppresses the single-marker flicker that a geometric prior
would otherwise emit as spurious one-marker segments.

```{r swap}
rigid <- fit(o, emission_count(err = 0.01),
             duration_rigidity(rigidity = 4L), priors = priors)
decode(rigid, o)
```

Or keep the duration and swap the **emission** to the categorical genotype model
(`emission_gt()`) when you have hard calls rather than read counts — the same axis
machinery, a different likelihood.

## Where the named callers fit

`caller_spec()` bundles these axis choices into the named presets, and
`call_states()` / `call_ancestry()` resolve a caller to its `(emission, duration,
priors)` and run the `fit`/`decode`/`to_segments` loop per (sample, chromosome)
for you:

```{r spec}
spec <- caller_spec("nnil", design = "BC2S2")
names(spec)
```

So the four grid cells are `nnil` = `emission_gt × duration_geometric`, `bbnil` =
`emission_count × duration_geometric`, `catiger` = `emission_gt ×
duration_rigidity`, and `rtiger` = `emission_count × duration_rigidity`. The
`binhmm`, `lbimpute`, and `fsfhap` callers carry extra
structure (binning, a coverage-aware emission, family pooling) and have their own
drivers, but they emit the same segment schema. See `vignette("callers")` for a
runnable example of each and `?caller_spec` for the axis defaults.
```
