---
title: "The engine: one chain, two swappable axes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The engine: one chain, two swappable axes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

`call_ancestry()` is a thin convenience layer. Underneath, every count/geometric
caller is the **same 3-state (REF/HET/ALT) HMM** parameterised along two
independent axes plus a set of design priors:

```
caller  =  emission model  ×  duration prior  +  design priors
```

- **Emission** — how an observation maps to a state likelihood:
  `emission_count()` (BetaBinomial over ref/alt depths) or `emission_gt()`
  (categorical genotype-error model).
- **Duration** — the transition prior on run lengths:
  `duration_geometric()`, `duration_rigidity()` (minimum run length), or
  `duration_hsmm()`.
- **Design priors** — `design_priors()` turns a breeding design into the state
  frequencies (`f_1`, `f_2`).

This vignette drives the engine directly with `fit()` → `decode()` →
`to_segments()`, which is exactly what `call_ancestry()` chains for you.

```{r setup}
library(nilHMM)
```

## The building blocks

Each axis is a small constructor returning a spec:

```{r blocks}
design_priors("BC2S2")     # state frequencies from the breeding design
emission_count(err = 0.01, conc = 20)
duration_geometric(rrate = 1e-4)
```

## Fit and decode by hand

`fit()` takes a per-(sample, chromosome) observation frame — `n` = depth, `a` =
alt-read count — an emission spec, a duration spec, and priors, and returns a
fitted model. `decode()` runs Viterbi and returns one **state per observation**
(coordinate-free: `0/1/2`).

```{r fit-decode}
# one sample, one chromosome: REF ... a short ALT run ... REF
obs <- data.frame(n = rep(8L, 12),
                  a = c(0, 1, 0, 0, 7, 8, 6, 7, 0, 1, 0, 0))
model <- fit(obs, emission_count(err = 0.01),
             duration_geometric(1e-4), priors = design_priors("BC2S2"))
decode(model, obs)
```

## Coordinates back on: `to_segments()`

`decode()` is coordinate-free. Attach `name/chr/pos` and run `to_segments()` to
collapse the per-marker states into the common segment schema (this is the second
half of `call_ancestry()`):

```{r segments}
states <- data.frame(name = "S1", chr = 1L,
                     pos = seq_len(nrow(obs)) * 1e5L,
                     state = decode(model, obs))
to_segments(states)
```

## Swapping an axis

Because the axes are independent, changing the caller's behaviour is a matter of
swapping one spec. Keep the same count emission but replace the geometric duration
with a **rigidity** prior (minimum run length 4) — the RTIGER-style segmentation —
and the shorter, noisier runs get absorbed:

```{r swap}
rigid <- fit(obs, emission_count(err = 0.01),
             duration_rigidity(rigidity = 4L), priors = design_priors("BC2S2"))
decode(rigid, obs)
```

Or keep the duration and swap the **emission** to the categorical genotype model
(`emission_gt()`) when you have hard calls rather than read counts — the same axis
machinery, a different likelihood.

## Where the named callers fit

`caller_spec()` bundles these axis choices into the named presets, and
`call_states()` / `call_ancestry()` resolve a caller to its `(emission, duration,
priors)` and run the `fit`/`decode`/`to_segments` loop per (sample, chromosome)
for you:

```{r spec}
spec <- caller_spec("nnil", design = "BC2S2")
names(spec)
```

So `nnil` is `emission_count × duration_geometric`, `rtiger` is
`emission_count × duration_rigidity`, and the genotype callers swap in
`emission_gt`. The `binhmm`, `lbimpute`, and `fsfhap` callers carry extra
structure (binning, a coverage-aware emission, family pooling) and have their own
drivers, but they emit the same segment schema. See `vignette("callers")` for a
runnable example of each and `?caller_spec` for the axis defaults.
```
