---
title: "Full-sib families with the fsfhap caller"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Full-sib families with the fsfhap caller}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

`fsfhap` is a native port of TASSEL's **FSFHap** (Swarts et al. 2014). Unlike the
per-line callers, it **pools each full-sib family**: the scattered introgressions
across the family's lines collectively phase the donor haplotype, then a 5-state
Viterbi-training EM imputes and smooths each line. It is bit-exact vs TASSEL on its
intended populations (backcross and RIL/inbred). It needs two things the per-line
callers don't: a called genotype `g`, and a **`family` grouping**.

```{r setup}
library(nilHMM)
```

## Input: HapMap + pedigree

FSFHap's native inputs are a TASSEL **HapMap** and a 7-column **pedigree**.
`read_hapmap()` and `read_pedigree()` normalise them to the engine's canonical
tables. Here we write tiny example files to demonstrate the readers (in practice
you point them at your files):

```{r io}
hmp <- tempfile(fileext = ".hmp.txt")
writeLines(c(
  "rs#\talleles\tchrom\tpos\tstrand\tassembly#\tcenter\tprotLSID\tassayLSID\tpanelLSID\tQCcode\tL1\tL2",
  "m1\tA/C\t1\t100\t+\tNA\tNA\tNA\tNA\tNA\tNA\tA\tC",
  "m2\tA/C\t1\t200\t+\tNA\tNA\tNA\tNA\tNA\tNA\tM\tA"), hmp)
read_hapmap(hmp)          # -> name, chr, pos, g   (g in {0 A-hom, 1 het, 2 C-hom, 3 NA})

ped_file <- tempfile()
writeLines(c("family\ttaxon\tparent1\tparent2\tp1\tp2\tF",
             "TIL01\tL1\tW22\tTIL01_teo\t0.75\t0.25\t0.9375",
             "TIL01\tL2\tW22\tTIL01_teo\t0.75\t0.25\t0.9375"), ped_file)
read_pedigree(ped_file)   # -> taxon, family, parents, contribution1/2, F
```

You attach the family grouping by joining on the taxon name:

```r
data <- read_hapmap("family.hmp.txt")
ped  <- read_pedigree("family_pedigree.txt")
data$family <- ped$family[match(data$name, ped$taxon)]
stopifnot(!anyNA(data$family))            # every sample must be in the pedigree
```

## Design routing derives `phet`

Pass a `design` token (`BC{n}S{m}`); the dispatcher selects the parent-calling
route and **derives the expected heterozygosity** `phet = (1 - F)/2` from the
design's inbreeding coefficient — no magic constant. BC1S4 (the TeoNAM design)
gives `F = 0.9375`, so `phet = 0.03125`. `fsfhap` supports BC1 designs
end-to-end; other designs route to as-yet-unported TASSEL paths.

## Calling on a family

A real family needs enough segregating markers to phase the donor (and to clear
FSFHap's coverage floor), so we simulate a 24-line BC1 family with donor blocks
summing to ~1/4 of the chromosome:

```{r family}
sim_family <- function(n_lines = 24L, m = 500L, donor_frac = 0.25, miss = 0.1) {
  pos <- as.integer(seq_len(m) * 1e5)
  do.call(rbind, lapply(seq_len(n_lines), function(l) {
    truth <- integer(m)
    for (b in seq_len(max(0L, rpois(1, donor_frac / 0.12)))) {
      w <- max(2L, rpois(1, m * 0.12)); s0 <- sample(m, 1L)
      truth[s0:min(m, s0 + w - 1L)] <- 2L
    }
    truth[truth == 0L & runif(m) < 0.03] <- 1L          # residual het
    g <- truth; g[runif(m) < miss] <- 3L                # skeleton missingness
    data.frame(name = sprintf("L%02d", l), family = "TIL01",
               chr = 1L, pos = pos, g = as.integer(g))
  }))
}
fam <- sim_family()

calls <- call_ancestry(fam, caller = "fsfhap", design = "BC1S4")
head(calls, 4)
table(calls$state)
```

## Parallelism

Each **family × chromosome** is an independent unit, so `fsfhap` fans them out
across cores with `threads` (bit-identical to serial). On a real full-genome
cohort this is a large win — the family × chromosome units fill all cores:

```{r threads}
identical(
  call_ancestry(fam, caller = "fsfhap", design = "BC1S4", threads = 1L),
  call_ancestry(fam, caller = "fsfhap", design = "BC1S4", threads = 2L)
)
```

On real TeoNAM (all five families, 51,004 markers, 50 units) the port runs in
**28.6 s on 10 threads vs TASSEL's 218.5 s** — 7.6× faster, and faster than TASSEL
even single-threaded. See `design/FSFHAP_PORT.md` for the full benchmark.

## When *not* to use `fsfhap`

`fsfhap` assumes homozygous founders and balanced full-sib segregation. For NILs
with **heterozygous or outbred donors** (e.g. teosinte), the two-haplotype model
breaks down — use the donor-agnostic `nnil` caller instead. Het-heavy F2 is a
documented stress case. See `vignette("callers")` for the alternatives and
`?call_ancestry` for the full parameter list.
```
