---
title: "Full-sib families with the fsfhap caller"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Full-sib families with the fsfhap caller}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_chunk$set(eval = requireNamespace("simcross", quietly = TRUE))
set.seed(1)
```

`fsfhap` is a native port of TASSEL's **FSFHap** (Swarts et al. 2014). Unlike the
per-line callers, it **pools each full-sib family**: the scattered introgressions
across the family's lines collectively phase the donor haplotype, then a 5-state
Viterbi-training EM imputes and smooths each line. It is bit-exact vs TASSEL on its
intended populations (backcross and RIL/inbred). It needs two things the per-line
callers don't: a called genotype `g`, and a **`family` grouping**.

```{r setup}
library(nilHMM)
```

## Input: HapMap + pedigree

FSFHap's native inputs are a TASSEL **HapMap** and a 7-column **pedigree**.
`read_hapmap()` and `read_pedigree()` normalise them to the engine's canonical
tables. Here we write tiny example files to demonstrate the readers (in practice
you point them at your files):

```{r io}
hmp <- tempfile(fileext = ".hmp.txt")
writeLines(c(
  "rs#\talleles\tchrom\tpos\tstrand\tassembly#\tcenter\tprotLSID\tassayLSID\tpanelLSID\tQCcode\tB1_L1\tB1_L2",
  "m1\tA/C\t1\t100\t+\tNA\tNA\tNA\tNA\tNA\tNA\tA\tC",
  "m2\tA/C\t1\t200\t+\tNA\tNA\tNA\tNA\tNA\tNA\tM\tA"), hmp)
read_hapmap(hmp)          # -> name, chr, pos, g   (g in {0 A-hom, 1 het, 2 C-hom, 3 NA})

# pedigree: family B1 = recurrent parent A x donor B1 (contribution 0.75 / 0.25)
ped_file <- tempfile()
writeLines(c("family\ttaxon\tparent1\tparent2\tp1\tp2\tF",
             "B1\tB1_L1\tA\tB1\t0.75\t0.25\t0.9375",
             "B1\tB1_L2\tA\tB1\t0.75\t0.25\t0.9375"), ped_file)
read_pedigree(ped_file)   # -> taxon, family, parents, contribution1/2, F
```

You attach the family grouping by joining on the taxon name:

```r
data <- read_hapmap("family.hmp.txt")
ped  <- read_pedigree("family_pedigree.txt")
data$family <- ped$family[match(data$name, ped$taxon)]
stopifnot(!anyNA(data$family))            # every sample must be in the pedigree
```

## Design routing derives `phet`

Pass a `design` token (`BC{n}S{m}`); the dispatcher selects the parent-calling
route and **derives the expected heterozygosity** `phet = (1 - F)/2` from the
design's inbreeding coefficient — no magic constant. BC1S4 (the TeoNAM design)
gives `F = 0.9375`, so `phet = 0.03125`. `fsfhap` supports BC1 designs
end-to-end; other designs route to as-yet-unported TASSEL paths.

## Calling on several families

FSFHap pools *each* family independently, so a realistic run has more than one —
each a distinct donor crossed onto the shared recurrent parent **A**. We simulate
three **BC1S4** families with [**simcross**](https://cran.r-project.org/package=simcross),
donors **B1**, **B2**, **B3** (the TeoNAM design: one backcross to A, then four
selfings — `F = 0.9375`, so `phet = 0.03125`). Each family needs enough
segregating markers to phase its donor and clear FSFHap's coverage floor, so we
use 20 lines × 500 markers. simcross gives real recombination breakpoints; the
donor blocks and residual heterozygosity fall out of the crossing scheme.

```{r family}
library(simcross)
sim_family <- function(donor, n_lines = 20L, m = 500L, L = 100, nself = 4L, miss = 0.1) {
  rec <- create_parent(L, 1); don <- create_parent(L, 2)  # recurrent A x this donor
  one_line <- function() {
    ind <- cross(cross(rec, don), rec)                  # F1 -> backcross to recurrent (BC1)
    for (s in seq_len(nself)) ind <- cross(ind, ind)    # four selfings -> BC1S4
    ind
  }
  map <- seq(0, L, length.out = m)
  g <- as.vector(convert2geno(lapply(seq_len(n_lines), function(i) one_line()), map)) - 1L
  g[runif(length(g)) < miss] <- 3L                      # skeleton missingness
  data.frame(name = rep(sprintf("%s_L%02d", donor, seq_len(n_lines)), times = m),
             donor = donor, family = donor, chr = 1L,
             pos = rep(as.integer(map * 1e6), each = n_lines), g = as.integer(g))
}
fam <- do.call(rbind, lapply(c("B1", "B2", "B3"), sim_family))

calls <- call_ancestry(fam, caller = "fsfhap", design = "BC1S4")
table(donor = calls$donor, state = calls$state)         # per-family state counts
head(calls, 4)
```

## Parallelism

Each **family × chromosome** is an independent unit — here three families on one
chromosome, so three units — and `fsfhap` fans them out across cores with
`threads` (bit-identical to serial):

```{r threads}
identical(
  call_ancestry(fam, caller = "fsfhap", design = "BC1S4", threads = 1L),
  call_ancestry(fam, caller = "fsfhap", design = "BC1S4", threads = 3L)
)
```

On a real full-genome cohort this is a large win. On real TeoNAM (all five
families, 51,004 markers, 50 family × chromosome units) the port runs in
**28.6 s on 10 threads vs TASSEL's 218.5 s** — 7.6× faster, and faster than TASSEL
even single-threaded. See `design/FSFHAP_PORT.md` for the full benchmark.

## When *not* to use `fsfhap`

`fsfhap` assumes homozygous founders and balanced full-sib segregation. For NILs
with **heterozygous or outbred donors** (e.g. teosinte), the two-haplotype model
breaks down — use the donor-agnostic `nnil` caller instead. Het-heavy F2 is a
documented stress case. See `vignette("callers")` for the alternatives and
`?call_ancestry` for the full parameter list.
```
