---
title: "Getting started with nilHMM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with nilHMM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 3)
# the data chunks use simcross (a Suggested package); skip gracefully if absent
knitr::opts_chunk$set(eval = requireNamespace("simcross", quietly = TRUE))
set.seed(1)
```

**nilHMM** calls ancestry — donor introgressions — in Near-Isogenic Lines and
related backcross/full-sib populations from sequencing data. Under the hood it is
a single **duration-aware 3-state (REF / HET / ALT) HMM engine** with two swappable
axes (emission × duration); their combinations, plus breeding-design priors,
express a family of named callers (the grid cells `nnil`, `bbnil`, `catiger`,
`rtiger`, plus `googa`, `atlas`, `binhmm`, `lbimpute`, `fsfhap`, `pedigree`). This
vignette covers the common workflow; see
`vignette("callers")` to choose a caller, `vignette("engine")` for the internals,
and `vignette("fsfhap")` for full-sib families.

```{r setup}
library(nilHMM)
```

## The one-verb API

Everything funnels through a single call:

```r
call_ancestry(data, caller = ..., design = ..., ...)
```

`data` is a tidy long table the readers produce; `caller` picks the model;
`design` supplies the breeding-design priors. The package is **data-agnostic** —
functions take `(data, params)` and never touch file paths, so you own where data
comes from and where results go.

## A worked example

We need an observation table. In practice you read one with `read_counts()` (a
`chr pos ref n_ref alt n_alt` TSV or a directory of them), `read_vcf_gt()`,
`read_hapmap()`, or `read_plink()`. Here we generate one with the package's own
design-driven simulator: `simulate_nil()` builds a breeding design and runs it
through [**simcross**](https://cran.r-project.org/package=simcross) (real meiosis
and recombination on the **bundled B73 v5 consensus map**, [load_map()]) to get
the true donor mosaic, and `simulate_counts()` degrades that truth to observed
allelic read counts. We take a **BC2S2** cohort — six lines on chromosomes 1–2,
donor **B** crossed onto recurrent parent **A** (so `REF` = A, `ALT` = donor B):

```{r simulate}
truth  <- simulate_nil("BC2S2", n = 6, chr = 1:2, n_markers = 300, donor = "B",
                       names = sprintf("NIL%02d", 1:6), seed = 1)   # true donor mosaic
counts <- simulate_counts(truth, depth = 6, seed = 1)[        # observed low-cov counts
  c("name", "donor", "chr", "pos", "n_ref", "n_alt")]
head(counts)
```

Call ancestry with the `bbnil` count caller and BC2S2 design priors:

```{r call}
calls <- call_ancestry(counts, caller = "bbnil", design = "BC2S2",
                       rrate = 1e-4, err = 0.01)
head(calls)
```

## The common segment schema

Every caller returns the same tidy **segment** table, so results are directly
comparable across callers and populations:

| column | meaning |
|---|---|
| `source` | free label for the run/method |
| `donor` | donor/taxon label (per line if the input carries one) |
| `name` | sample identifier |
| `chr` | chromosome |
| `start_bp`, `end_bp` | segment bounds (bp) |
| `state` | `REF` (0, recurrent hom), `HET` (1), or `ALT` (2, donor hom) |

```{r schema}
table(calls$state)
```

Numeric states are `0 = REF`, `1 = HET`, `2 = ALT`. The package's `paint_calls()`
renders those segments as a chromosome painting — one band per line, faceted by
chromosome (REF gold / HET green / ALT purple):

```{r plot, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("simcross", quietly = TRUE), fig.width = 7, fig.height = 3.2}
paint_calls(calls)
```

## Which caller?

All callers share the REF/HET/ALT chain and the design priors; they differ in the
emission model, the duration prior, and the input they expect.

| caller | input | when |
|---|---|---|
| `nnil` | called `GT` (hard genotypes) | categorical (gt) + geometric; Holland's nNIL on hard calls |
| `bbnil` | allelic read counts | count/BetaBinomial + geometric; low-coverage skim/BrB |
| `catiger` | called `GT` (hard genotypes) | categorical (gt) + rigidity |
| `rtiger` | allelic read counts | count + minimum-run-length (rigidity) segmentation |
| `binhmm` | allelic read counts | per-bin calling for noisy/uneven coverage |
| `googa` | recurrent/donor counts | competitive-alignment RNA-seq (faithful GOOGA; gt + geometric) |
| `atlas` | recurrent/donor counts | competitive-alignment RNA-seq (this work; gt + rigidity) |
| `lbimpute` | allelic read counts | biallelic imputation (LB-Impute port; distance-dependent transition) |
| `fsfhap` | called `GT` + a `family` | full-sib families (TASSEL FSFHap) |
| `pedigree` | counts or hard-call `state`/`g` + a pedigree | family-coupled belief propagation |

`vignette("callers")` gives runnable examples for the per-sample callers (see
`vignette("fsfhap")` for the family workflow). The no-HMM per-site *genotype*
baseline (the het-excess "control") is `call_gt()`, not an ancestry caller — see
`?call_gt`.

## Next steps

- **Choosing and comparing callers:** `vignette("callers")`
- **The engine (emission × duration, custom callers):** `vignette("engine")`
- **Full-sib families (HapMap/pedigree → fsfhap):** `vignette("fsfhap")`
- Function reference: `?call_ancestry`, `?read_counts`, `?to_segments`.
```
