---
title: "Getting started with nilHMM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with nilHMM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 3)
set.seed(1)
```

**nilHMM** calls ancestry — donor introgressions — in Near-Isogenic Lines and
related backcross/full-sib populations from sequencing data. Under the hood it is
a single **duration-aware 3-state (REF / HET / ALT) HMM engine** with two swappable
axes (emission × duration); their combinations, plus breeding-design priors,
express a family of named callers (`nnil`, `rtiger`, `binhmm`, `atlas`,
`lbimpute`, `fsfhap`). This vignette covers the common workflow; see
`vignette("callers")` to choose a caller, `vignette("engine")` for the internals,
and `vignette("fsfhap")` for full-sib families.

```{r setup}
library(nilHMM)
```

## The one-verb API

Everything funnels through a single call:

```r
call_ancestry(data, caller = ..., design = ..., ...)
```

`data` is a tidy long table the readers produce; `caller` picks the model;
`design` supplies the breeding-design priors. The package is **data-agnostic** —
functions take `(data, params)` and never touch file paths, so you own where data
comes from and where results go.

## A worked example

We need an observation table. In practice you read one with `read_counts()` (a
`chr pos ref n_ref alt n_alt` TSV or a directory of them), `read_vcf_gt()`,
`read_hapmap()`, or `read_plink()`. Here we simulate a small cohort of allelic
read counts so the vignette is self-contained: six lines, two chromosomes, each
carrying one donor (ALT) introgression block on a REF background.

```{r simulate}
sim_counts <- function(n = 6L, m = 300L, n_chr = 2L, depth = 6, err = 0.01) {
  chr <- rep(seq_len(n_chr), each = m)
  pos <- as.integer(rep(seq_len(m), n_chr) * 1e5)
  do.call(rbind, lapply(seq_len(n), function(s) {
    truth <- integer(length(chr))                        # 0 REF / 1 HET / 2 ALT
    for (ch in seq_len(n_chr)) {
      idx <- which(chr == ch); w <- sample(20:60, 1L); s0 <- sample(idx, 1L)
      blk <- s0:min(max(idx), s0 + w); truth[blk] <- 2L; truth[blk[1]] <- 1L
    }
    d  <- rpois(length(chr), depth)
    na <- rbinom(length(chr), d, c(err, 0.5, 1 - err)[truth + 1L])
    data.frame(name = sprintf("NIL%02d", s), chr = chr, pos = pos,
               n_ref = d - na, n_alt = na)
  }))
}
counts <- sim_counts()
head(counts)
```

Call ancestry with the `nnil` count caller and BC2S2 design priors:

```{r call}
calls <- call_ancestry(counts, caller = "nnil", design = "BC2S2",
                       rrate = 1e-4, err = 0.01)
head(calls)
```

## The common segment schema

Every caller returns the same tidy **segment** table, so results are directly
comparable across callers and populations:

| column | meaning |
|---|---|
| `source` | free label for the run/method |
| `donor` | donor/taxon label (per line if the input carries one) |
| `name` | sample identifier |
| `chr` | chromosome |
| `start_bp`, `end_bp` | segment bounds (bp) |
| `state` | `REF` (0, recurrent hom), `HET` (1), or `ALT` (2, donor hom) |

```{r schema}
table(calls$state)
```

Numeric states are `0 = REF`, `1 = HET`, `2 = ALT`. A quick karyotype-style view
of one line's introgressions (base graphics, no extra packages):

```{r plot}
one <- calls[calls$name == "NIL01", ]
cols <- c("0" = "grey85", "1" = "goldenrod", "2" = "firebrick")
plot(NA, xlim = c(0, max(one$end_bp)), ylim = c(0.5, max(one$chr) + 0.5),
     xlab = "position (bp)", ylab = "chromosome", yaxt = "n",
     main = "NIL01 ancestry")
axis(2, at = sort(unique(one$chr)))
with(one, rect(start_bp, chr - 0.3, end_bp, chr + 0.3,
               col = cols[as.character(state)], border = NA))
legend("topright", fill = cols, legend = c("REF", "HET", "ALT"), bty = "n")
```

## Which caller?

All callers share the REF/HET/ALT chain and the design priors; they differ in the
emission model, the duration prior, and the input they expect.

| caller | input | when |
|---|---|---|
| `nnil` | allelic counts, or called `GT` | general-purpose NILs (low-cov skim or saturated GT) |
| `rtiger` | allelic counts | minimum-run-length (rigidity) segmentation |
| `binhmm` | allelic counts | per-bin calling for noisy/uneven coverage |
| `atlas` | recurrent/donor counts | competitive-alignment RNA-seq (GOOGA) |
| `lbimpute` | very low-cov counts (<1×) | GBS/skim imputation (LB-Impute) |
| `fsfhap` | called `GT` + a `family` | full-sib families (TASSEL FSFHap) |

`vignette("callers")` gives a runnable example for each.

## Next steps

- **Choosing and comparing callers:** `vignette("callers")`
- **The engine (emission × duration, custom callers):** `vignette("engine")`
- **Full-sib families (HapMap/pedigree → fsfhap):** `vignette("fsfhap")`
- Function reference: `?call_ancestry`, `?read_counts`, `?to_segments`.
```
